Higher-Order unification for free!

### 2024-04-12 paper/main.tex
finish intro

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three designated type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\sigma}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of instances as rules (a.k.a. clauses).
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets delimit a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

@@

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

\begin{elpicode}
decision (all (app [con"fin", con"7"]) y\              ~\customlabel{goal:g}{(g)}~
  app [con"nfact", y, con"3"]).
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", y, con"3"] = app [P, y]               ~\customlabel{problem:p}{(p)}~
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", y, con"3"] = Pm y                    ~\customlabel{problem:pa}{(p')}~
Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\rho)}~
\end{elpicode}

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.
In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}). We show below the premise before and
after the instantiation of \elpiIn{P}:

\begin{elpicode}
decision (app [                   P                  , w])
decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
\end{elpicode}

\noindent
In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
the following unification problem has no solution:

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, that in turns makes the 
unification procedures of the meta language weak.\\
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons automatic proof search procedure typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs some care in order to
simulate a logic program in CIC using the unification of Elpi.

% For example:\\
% %
% \vspace{4pt}
% {
% \setlength{\tabcolsep}{2pt}
% \begin{tabular}{lcl|}
% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
% \end{tabular}
% \phantom{mn}
% \begin{tabular}{lcl}
% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
% \end{tabular}
% }
% \vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\todo{fix}\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:grounwork} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

The $\lambda$Prolog code discussed in the paper can be accessed at the
URL: \url{https://github.com/FissoreD/paper-ho}.

@@

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a rich(er) equational
theory and a \Ho{} meta language with a simpler one, and we reuse
the unification procedure of \Ho{} in order to implement one for \Fo.

>>>>

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012}
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, while square brackets delimit a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

@@

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

\begin{elpicode}
decision (all (app [con"fin", con"7"]) x\              ~\customlabel{goal:g}{(g)}~
  app [con"nfact", x, con"3"]).
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", x, con"3"] = app [P, x]               ~\customlabel{problem:p}{(p)}~
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", x, con"3"] = Pm x                    ~\customlabel{problem:pa}{(p')}~
Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\sigma)}~
\end{elpicode}

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.
In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}).
% We show below the premise before and
% after the instantiation of \elpiIn{P}:

% \begin{elpicode}
% decision (app [                   P                  , w])
% decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
% \end{elpicode}

% \noindent
In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
the following unification problem has no solution:

\begin{elpicode}
app [ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app [ con"nfact"                             , N, NF]
\end{elpicode}

\noindent
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, that in turns makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons automatic proof search procedure typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs some care in order to
simulate a logic program in CIC using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta-language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \url{https://github.com/FissoreD/paper-ho}.

@@

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a richer equational
theory and a \Ho{} meta language with a simpler one.
